---
title: "Prosper Loan Data"
author: "Kevin Green"
date: "August 4th, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(scales)
library(ggplot2)
library(dplyr)
library(reshape2)
loans <- read.csv('prosperLoanData.csv')

```

<h2>Abstract</h2>
I describe the effects that certain factors play over
<h2>Introduction</h2>
This data set contains information about loans. Has over 113000+ observations and 81 variables. Variables include data about the loan, borrower, lenders and investors. Studying this data should help understand factors that have an effect with loan agreements. For my personal benefit, I hope to better understand the factors that could help me better obtain a comfortable mortage and pay it off to finally own a house after wishing it for years.


<h2>Slight adjustment to the data sets</h2>
Started by removing ambigous employment statuses from the data set also removed outliers which made the charts very difficult to read. In addition, created a data frame with means and medians to better compare this information.

```{r}
#Subset data to exclude ambiguous employment status, top 1% earners 
#and bottom 1% among other outliers
loans2 <- subset(loans, !(EmploymentStatus %in% c('','Not available','Other')))
loans2 <- subset(loans2, StatedMonthlyIncome<quantile(StatedMonthlyIncome, 0.99) 
                 & StatedMonthlyIncome>quantile(StatedMonthlyIncome, 0.01) )
loans2 <- subset(loans2, !(IncomeRange %in% c('$0','Not displayed')))
loans2 <- subset(loans2, BankcardUtilization<quantile(BankcardUtilization,0.99, na.rm = TRUE) & CurrentDelinquencies<quantile(CurrentDelinquencies,0.99,na.rm = TRUE))

#Creating data frames with means and medians for Loan Amounts
loan.loan_by_income <- loans2 %>%
  group_by(IncomeRange) %>%
  summarise(mean_loan_amount=mean(LoanOriginalAmount),
            median_loan_amount=median(LoanOriginalAmount))

loan.loan_by_term <- loans2 %>%
  group_by(Term) %>%
  summarise(mean_loan_amount=mean(LoanOriginalAmount),
            median_loan_amount=median(LoanOriginalAmount))

mdata <- melt(loan.loan_by_income, id = c('IncomeRange'))
mdata2 <- melt(loan.loan_by_term, id = c('Term'))
```

<h2>Univariate Analysis</h2>
```{r}
#Chart 1
#This chart shows Employed status as the most common status, with 65998 instances
#Counts for Employment Status
ggplot(loans2, aes(EmploymentStatus)) +
geom_bar() +
geom_text(stat='count', aes(label=..count..), vjust=-0.5)+ xlab('Employment Status') + ylab('Count')

by(loans2$StatedMonthlyIncome, loans2$IncomeRange, summary)
#Chart 3
#Loans by Income range
ggplot(loans2, aes(IncomeRange))+
  geom_bar()+xlab('Income Range')+ylab('Count')

#Chart 6
#Loans defaulted by IncomeRange
ggplot(subset(loans2, LoanStatus == 'Defaulted'), aes(IncomeRange)) +
geom_histogram(stat = 'count')+
  scale_x_discrete(limits=c('Not employed','$1-24,999','$25,000-49,999','$50,000-74,999',
                            '$75,000-99,999','$100,000+'))+xlab('Income Range')+ylab('Count')

#Chart 7
#Loans Completed by IncomeRange
ggplot(subset(loans2, LoanStatus == 'Completed'), aes(IncomeRange)) +
geom_histogram(stat = 'count')+
  scale_x_discrete(limits=c('Not employed','$1-24,999','$25,000-49,999','$50,000-74,999',
                            '$75,000-99,999','$100,000+'))+xlab('Income Range')+ylab('Count')

#Chart 9
#Counts for IsBorrowerHomeowner
ggplot(loans2, aes(IsBorrowerHomeowner)) +
geom_bar() +
geom_text(stat='count', aes(label=..count..), vjust=-0.5)+xlab('Borrower is Homeowner') + ylab('Count')

#Chart 13
#Loan amounts
ggplot(loans2, aes(LoanOriginalAmount)) +
geom_histogram()+xlab('Loan Amount') + ylab('Count')

#Chart 14
#Loans defaulted by Original Amount
ggplot(subset(loans2, LoanStatus == 'Defaulted'), aes(LoanOriginalAmount)) +
geom_histogram()+xlab('Loan Amount') + ylab('Count')

#Chart 15
#Loans completed by Original Amount
ggplot(subset(loans2, LoanStatus == 'Completed'), aes(LoanOriginalAmount)) +
geom_histogram()+xlab('Loan Amount')+ylab('Count')
```

<h2>Bivariate Analysis</h2>

```{r}
#Chart 2
#As well this charts shows Employed status
#LoanOriginalAmount by EmploymentStatus 
ggplot(loans2, aes(EmploymentStatus, LoanOriginalAmount))+
  geom_boxplot()+ xlab('Employment Status') + ylab('Loan Amount')


#Chart 4
#LoanOriginalAmount by IncomeRange 
ggplot(loans2, aes(IncomeRange, LoanOriginalAmount))+
  geom_boxplot()+ xlab('Income Range') + ylab('Loan Amount')+
  scale_x_discrete(limits=c('$1-24,999','$25,000-49,999','$50,000-74,999',
                            '$75,000-99,999','$100,000+'))
#Chart 5
#DebtToIncomeRatio for IncomeRange
ggplot(loans2, aes(x = IncomeRange, y =DebtToIncomeRatio))+
  geom_bar(stat = 'summary', fun.y = mean)+
  scale_x_discrete(limits=c('$1-24,999','$25,000-49,999','$50,000-74,999',
                            '$75,000-99,999','$100,000+'))+
  facet_wrap(~IsBorrowerHomeowner)+xlab('Income Range')+ylab('Debt to Income Ratio')

#Chart 8
#Mean and median loan amount by IncomeRange
ggplot(mdata, aes(x = IncomeRange, value, fill = variable))+
  geom_bar(position = 'dodge', stat = 'identity')+
  scale_x_discrete(limits=c('Not employed','$1-24,999','$25,000-49,999','$50,000-74,999',
                            '$75,000-99,999','$100,000+'))+xlab('Income Range')+ylab('Mean and Median')

#Chart 10
#Mean StatedMonthlyIncome for IsBorrowerHomeowner
ggplot(loans2, aes(IsBorrowerHomeowner, StatedMonthlyIncome))+
  geom_bar(stat = 'summary', fun.y = mean)+xlab('Borrower is Homeowner') + ylab('Monthly Income')

#Chart 11
#LoanOriginalAmount by IsBorrowerHomeowner (EmploymentStatus)
ggplot(loans2, aes(IsBorrowerHomeowner, LoanOriginalAmount))+
  geom_boxplot()+ xlab('Homeowner') + ylab('Loan Amount')+
  facet_wrap(~EmploymentStatus)

#Chart 12
#LoanOriginalAmount by IsBorrowerHomeowner (IncomeRange)
ggplot(loans2, aes(IsBorrowerHomeowner, LoanOriginalAmount))+
  geom_boxplot()+ xlab('Homeowner') + ylab('Loan Amount')+
  facet_wrap(~IncomeRange)

#Chart 16
#Mean and median  amount by Term
ggplot(mdata2, aes(Term, value, fill = variable))+
  geom_bar(position = 'dodge', stat = 'identity')+xlab('Term')+ylab('Mean and Median')

#Chart 17
#LoanAmount by MonthlyIncome
ggplot(loans2, aes(StatedMonthlyIncome, LoanOriginalAmount))+
  geom_point(alpha  = 0.05, position=position_jitter())+xlab('Monthly Income')+ylab('Loan Amount')

#Chart 20
#Loan amount by credit score
ggplot(loans2,aes(CreditScoreRangeLower, LoanOriginalAmount))+
  geom_point(alpha = 0.10)+
  geom_smooth(method='auto', color='red')+xlab('Credit Score Lower Range')+ylab('Loan Amount')

#Chart 21
#Rate by credit score
ggplot(loans2,aes(CreditScoreRangeLower, BorrowerRate))+
  geom_point(alpha = 0.10)+
  geom_smooth(method='auto', color='red')+xlab('Credit Score Lower Range')+ylab('Borrower Rate')

#Chart 22
#APR by credit score
ggplot(loans2,aes(CreditScoreRangeLower, BorrowerAPR))+
  geom_point(alpha = 0.10)+
  geom_smooth(method='auto', color='red')+xlab('Credit Score Lower Range')+ylab('Borrower APR')

#Chart 23
#LenderYield by BorrowerAPR
ggplot(loans2, aes(BorrowerAPR, LenderYield))+
  geom_point(alpha = 0.10)+xlab('Borrower APR')+ylab('Lender Yield')

cor.test(loans2$BorrowerAPR, loans2$LenderYield, method='pearson')
#cor = 0.9910457 | Very strong positive relation

#Chart 24
#EstimatedLoss by BorrowerAPR
ggplot(loans2, aes(BorrowerAPR, EstimatedLoss))+
  geom_point(alpha = 0.10)+xlab('Borrower APR')+ylab('Estimated Loss')

#Chart 25
#LenderYield by EstimatedLoss
ggplot(loans2, aes(LenderYield, EstimatedLoss))+
  geom_point(alpha = 0.10)+xlab('Lender Yield')+ylab('Estimated Loss')

#Chart 26
#LenderYield by ProsperScore
ggplot(loans2, aes(x = ProsperScore, y = LenderYield)) +
  geom_line(stat='summary', fun.y=median)+xlab('Prosper Score')+ylab('Lender Yield')

#Chart 29
#StatedMonthlyIncome by EmploymentStatusDuration
ggplot(loans2, aes(x = EmploymentStatusDuration, y = StatedMonthlyIncome))+
  geom_line(stat='summary', fun.y=median)+xlab('Employment Status Duration')+ylab('Monthly Income')

#Chart 30
#StatedMonthlyIncome by OpenCreditLines
ggplot(subset(loans2, OpenCreditLines<quantile(OpenCreditLines, 0.99,  na.rm=TRUE)), aes(x = OpenCreditLines, y = StatedMonthlyIncome))+
  geom_line(stat='summary', fun.y=mean)+xlab('Open Credit Lines')+ylab('Monthly Income')
cor.test(loans2$OpenCreditLines, loans2$StatedMonthlyIncome, method='pearson')
# cor = 0.2826184 | Not a very strong relationship

#Chart 31
#StatedMonthlyIncome by InquiriesLast6Months
ggplot(subset(loans2, InquiriesLast6Months<quantile(InquiriesLast6Months, 0.99,  na.rm=TRUE)), aes(x = InquiriesLast6Months, y = StatedMonthlyIncome))+
  geom_line(stat='summary', fun.y=mean)+xlab('Inquiries in the Last 6 Months')+ylab('Monthly Income')

#Chart 32
#StatedMonthlyIncome by CurrentDelinquencies
ggplot(subset(loans2, CurrentDelinquencies<quantile(CurrentDelinquencies, 0.99,  na.rm=TRUE)), aes(x = CurrentDelinquencies, y = StatedMonthlyIncome))+
  geom_line(stat='summary', fun.y=mean)+xlab('Current Delinquencies')+ylab('Monthly Income')

#Chart 33
#InquiriesLast6Months by CurrentDelinquencies
ggplot(subset(loans2, CurrentDelinquencies<quantile(CurrentDelinquencies, 0.99,  na.rm=TRUE)), aes(x = CurrentDelinquencies, y = InquiriesLast6Months))+
  geom_line(stat='summary', fun.y=mean)+xlab('Current Delinquencies')+ylab('Inquiries in the Last 6 Months')

#Chart 34
#CurrentCreditLines by CurrentDelinquencies
ggplot(subset(loans2, CurrentDelinquencies<quantile(CurrentDelinquencies, 0.99,  na.rm=TRUE)), aes(x = CurrentDelinquencies, y = CurrentCreditLines))+
  geom_line(stat='summary', fun.y=mean)+xlab('Current Delinquencies')+ylab('Number of Credit Lines')

#Chart 35
#LoanOriginalAmount by CurrentDelinquencies
ggplot(subset(loans2, CurrentDelinquencies<quantile(CurrentDelinquencies, 0.99,  na.rm=TRUE)), aes(x = CurrentDelinquencies, y = LoanOriginalAmount))+
  geom_line(stat='summary', fun.y=mean)+xlab('Current Delinquencies')+ylab('Loan Amount')

```



<h2>Multivariate Analysis</h2> 

```{r}
#Chart 18
#LoanAmount by MonthlyIncome (Colored)
ggplot(loans2, aes(StatedMonthlyIncome, LoanOriginalAmount, color = IncomeRange))+
  geom_point(position=position_jitter())+xlab('Monthly Income')+ylab('Loan Amount')

#Chart 19
#LoanAmount by MonthlyIncome (Term)
ggplot(loans2, aes(StatedMonthlyIncome, LoanOriginalAmount))+
  geom_point(alpha = 0.3, position=position_jitter())+
  facet_wrap(~Term)+xlab('Monthly Income')+ylab('Loan Amount')

#Chart 27
#LenderYield by ProsperScore (IncomeRange)
ggplot(loans2, aes(x = ProsperScore, y = LenderYield, color = IncomeRange)) +
  geom_line(stat='summary', fun.y=median)+xlab('Prosper Score')+ylab('Lender Yield')

#Chart 28
#BorrowerAPR by ProsperScore (IncomeRange)
ggplot(loans2, aes(x = ProsperScore, y = BorrowerAPR, color = IncomeRange)) +
  geom_line(stat='summary', fun.y=median)+xlab('Prosper Score')+ylab('Borrower APR')
```

<h2>Final Plots and Summary</h2>
The most interesting observations I found among this data, was not the relations between variables, but instead the lack of relation between variables where I was expecting the opposite.
One thing I was curious about was seeing which credit score were defaulting the most, before plotting the chart, I pictured lower credit scores defaulting the most, however, since most of the loans are given to borrowers with scores around the 700 vecinity, these scores were also the ones reporting most default loans likely due to the portion of the borrowers they represent. 

```{r}
#Default by Credit Score
ggplot(subset(loans2, LoanStatus == 'Defaulted'), aes(CreditScoreRangeLower))+
  geom_histogram(aes(y = (..count..)/sum(..count..)), binwidth=50)+
  scale_y_continuous(labels = scales::percent)+
  scale_x_continuous(limits = c(450, 900), breaks = seq(450, 900, 50))+xlab('Credit Score')+ylab('Percent Defaulted')+
  ggtitle('Default Distribution by Credit Score')
```

```{r}
#Default by Credit Score and Monthly Income
ggplot(subset(loans2, LoanStatus == 'Defaulted'),aes(CreditScoreRangeLower, StatedMonthlyIncome))+
  geom_line(stat = 'summary', fun.y = mean)+geom_smooth(method='auto', color='red')+
   ggtitle('Income by Credit Score Where Laons Defaulted')+xlab('Credit Score')+ylab('Monthly Income')
cor.test(loans2$CreditScoreRangeLower, loans2$StatedMonthlyIncome, method='pearson')
#cor = 0.1983418  | None to weak relationship
```
One thing I was very confident about was the fact that people with higher bankcard utlization would be more likely fail to pay on time. But my assumption was wrong again. In fact, this is one of the weakest relationships I explored, almost 0. 
```{r}
#Relationship between BankcardUtilization and CurrentDelinquencies
ggplot(loans2, aes(CurrentDelinquencies, BankcardUtilization))+
  geom_point(alpha=1/10, position = position_jitter(h=0))+
  ggtitle('Income by Credit Score Where Laons Defaulted')+
  xlab('Current Deliquencies')+ylab('Bankcard Utilization')+
  scale_x_continuous(limits = c(1,6), breaks = seq(1,6,1))+
  scale_y_continuous(limits = c(0.01,1))

cor.test(loans2$BankcardUtilization, loans2$CurrentDelinquencies, method='pearson')
#-0.0431209 | No relationship
````
<h2>Reflection</h2>
When I moved into this country, I had to build a credit score, given the fact that I had none. It was a struggle, because no financial institution would even let me open a credit card with them and how are you supposed to build a credit score if no one gives you the opportunity to build credit? Well, you start with a secured credit card, which essentially is paying the bank fees and interest for you to borrow from your own money, but these payments are reported to credit bureaus and that's how you start buidling credit history.
I knew that I would give my 100% to the bank to pay them back, but me telling them did not mean anything, because they really didn't know anything about me. Data speaks for itself.

Banks have performed these type of analysis thousands and thousands of times, likely with much more depth than this, so when they ask you about your credit score, your current debt, income and more, it is for a reason. Although you may think, these factors do not apply to you, because you know you will pay back, the bank has no way of measuring your ability to pay by just listening to you say so. In a greater sense, these factors are acuarate and help minimize the losses to both borrowers and lenders.

From a technical perspective, most of the challenges I faced while building these plots came from understanding or finding a plot that depicts data that makes sense. I had the variables and knew what I was looking to see, however building was hard, I either had line graphs with way too much noise, irrational bar graphs or scatter plots with lines of dots that made no sense. I also wanted to build some pie charts, and I can build simple ones, but was not able to build one with the data from this set, as I lacked the knowledge to do so, even after hours of researching online how to possibly do this.







